---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Introduction

### Summary
In this document the author presents an exploratory analysis of the publicly available dataset about Behavioral Risk Factor Surveillance System (<a href="http://www.cdc.gov/brfss/">BRFSS</a>).


### Abstract/General information about the dataset 

<i>"The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative project between all of the states in the United States (US) and participating US territories and the Centers for Disease Control and Prevention (CDC). The BRFSS is administered and supported by CDC's Population Health Surveillance Branch, under the Division of Population Health at the National Center for Chronic Disease Prevention and Health Promotion. BRFSS is an ongoing surveillance system designed to measure behavioral risk factors for the non-institutionalized adult population (18 years of age and older) residing in the US. The BRFSS was initiated in 1984, with 15 states collecting surveillance data on risk behaviors through monthly telephone interviews. Over time, the number of states participating in the survey increased; by 2001, 50 states, the District of Columbia, Puerto Rico, Guam, and the US Virgin Islands were participating in the BRFSS. Today, all 50 states, the District of Columbia, Puerto Rico, and Guam collect data annually and American Samoa, Federated States of Micronesia, and Palau collect survey data over a limited point- in-time (usually one to three months). In this document, the term “state” is used to refer to all areas participating in BRFSS, including the District of Columbia, Guam, and the Commonwealth of Puerto Rico.

The BRFSS objective is to collect uniform, state-specific data on preventive health practices and risk behaviors that are linked to chronic diseases, injuries, and preventable infectious diseases that affect the adult population. Factors assessed by the BRFSS in 2013 include tobacco use, HIV/AIDS knowledge and prevention, exercise, immunization, health status, healthy days — health-related quality of life, health care access, inadequate sleep, hypertension awareness, cholesterol awareness, chronic health conditions, alcohol consumption, fruits and vegetables consumption, arthritis burden, and seatbelt use. Since 2011, BRFSS conducts both landline telephone- and cellular telephone-based surveys. In conducting the BRFSS landline telephone survey, interviewers collect data from a randomly selected adult in a household. In conducting the cellular telephone version of the BRFSS questionnaire, interviewers collect data from an adult who participates by using a cellular telephone and resides in a private residence or college housing.

Health characteristics estimated from the BRFSS pertain to the non-institutionalized adult population, aged 18 years or older, who reside in the US. In 2013, additional question sets were included as optional modules to provide a measure for several childhood health and wellness indicators, including asthma prevalence for people aged 17 years or younger."
</i>
Source: <a href="http://guides.library.duke.edu/c.php?g=289704&p=1930838">Duke University Data and Visualization Services</a>


## Reproduceability - local setup

### Load packages

The current document makes use of these packages: 
```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
options(dplyr.width = Inf)
```

### Load data

In this repository, the data file is located in a dedicated directory called 'data'. 
Thus, it requires some "extra" effort to load it, which is NOT necessary in case you decide to co-locate the dataset to this R mark-down file.  

To load the dataset to the current working environment, you can do as follows: 

```{r load-data}
# Set your working directory to this repo's root BEFORE running next commands; for example, like so: setwd("./dir1/dir2/")
script.dir = getwd() 
data.dir = paste(sep="/", script.dir, "data")
data.file = paste(sep="/", data.dir, "brfss2013.RData")
load(data.file)
```
This will store in your global environment dataframe called "brfss2013".

The dataset contains an extensive amount of variables (330), and observations (491775): 

```{r dimensions-dataset}
dim(brfss2013)
```

To get more meta info about the dataset:

```{r summary-overview}
str(brfss2013)
summary(brfss2013)
```


* * *

## Part 1: About Data collection

Prior to analysing the data, it is essencial that one investigates how the sample data was collected, in order to know its limitations and potential in terms of inference. Namely, we want to know if it is generalizable and if we can prove causality.

The type of this study is "observational", as oposed to "experiment", since it consists of survey collection via telephone interviews.

The telephone interviews are conducted through random sampling from adult (aged 18 years or older) US population, as mentioned in the brfss_codebook: "<i>... interviewers collect data from a randomly selected adult in a household.</i>".
Moreover it is mentioned that "<i>..., interviewers collect data from an adult who participates by using a cellular telephone and resides in a private residence or college housing.</i>". S
ince the data collected depends on voluntary response, there might be an inherent bias involved. For example, some individuals who suffer from a given desease/health condition might be more reluctant to participate in such a survey.

Nonetheless, it is assumed that the sample is representative of the population at hand, given its large sample size (491775 observations) and randomness.

```{r number-of-observations}
dim(brfss2013)[2]
```

We can thus conclude that since this is an observational study which used random sampling, researchers can at best generalize conclusions, although not prove causality. This is due to the fact that no random assignment was used while conducting the study. 


* * *

## Part 2: Research questions

**Research quesion 1:** Regarding "Mental Illness and Stigma" and "Social Context", does the respondants' frequency of feeling of depression in the last 30 days vary according to the number of hours they work per week?

- Explanatory variable: "scntlwk1" (How Many Hours Per Week Did You Work); variable type: numerical discrete; data type encoding in dataset: int
- Response variable: "misdeprd" (How Often Feel Depressed Past 30 Days); variable type: categorical ordinal with 5 levels

**Research quesion 2:** Regarding "Healthy Days (Symptoms)" and "Demographics", does the respondants' energy level self-perception vary according to level of income?

- Response variable: "income2" (Income Level); variable type: categorical ordinal with 8 levels;
- Response variable: "qlhlth2" (How Many Days Full Of Energy In Past 30 Days); variable type: numerical discrete; data type encoding in dataset: int


**Research quesion 3:** Regarding "Social Context" and "Demographics", does the respondants' level of income vary according to the number of hours worked?

- Explanatory variable: "scntlwk1" (How Many Hours Per Week Did You Work); variable type: numerical discrete; data type encoding in dataset: int
- Response variable: "income2" (Income Level); variable type: categorical ordinal with 8 levels;

* * *

## Part 3: Exploratory data analysis

**Research quesion 1:** Regarding "Mental Illness and Stigma" and "Social Context", does the respondants' frequency of feeling of depression in the last 30 days vary according to the number of hours they work per week?

- Explanatory variable: "scntlwk1" (How Many Hours Per Week Did You Work); variable type: numerical discrete; data type encoding in dataset: int
- Response variable: "misdeprd" (How Often Feel Depressed Past 30 Days); variable type: categorical ordinal with 5 levels

Let us start by analysing the different levels of depression assessment.
```{r}
table(brfss2013$misdeprd)
plot(brfss2013$misdeprd)
```

Number of hours worked vs depression level
```{r}
table(brfss2013$scntlwk1,brfss2013$misdeprd)
plot(brfss2013$misdeprd, brfss2013$scntlwk1)
```

**Research quesion 2:** Regarding "Healthy Days (Symptoms)" and "Demographics", does the respondants' energy level self-perception vary according to level of income?

- Response variable: "income2" (Income Level); variable type: categorical ordinal with 8 levels;
- Response variable: "qlhlth2" (How Many Days Full Of Energy In Past 30 Days); variable type: numerical discrete; data type encoding in dataset: int


```{r}
table(brfss2013$scntvot1,brfss2013$misdeprd)
plot(brfss2013$misdeprd, brfss2013$scntvot1)
```


**Research quesion 3:** Regarding "Social Context" and "Demographics", does the respondants' level of income vary according to the number of hours worked?

- Explanatory variable: "scntlwk1" (How Many Hours Per Week Did You Work); variable type: numerical discrete; data type encoding in dataset: int
- Response variable: "income2" (Income Level); variable type: categorical ordinal with 8 levels;


```{r}
table(brfss2013$income2,brfss2013$misdeprd)
plot(brfss2013$misdeprd, brfss2013$income2)
```
